Linting Linearity in Core/System FC

### 2022-11-25 proposal.tex
tweaks

Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types to a mainstream, pure, and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once (its arguments have multiplicity of one), for some
definition of \emph{consume} that I should revise here.

@@

\subsection{Handling the case binder\label{casebinder}}

@@

The definition for $\$!$ in \textbf{linear-base}\cite{linearbase} fails to lint after the
desugarer (before any optimisation takes place) with \emph{Linearity failure in
lambda: x 'Many $\not\subseteq$ p}
\begin{code}
($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
($!) f !x = f x
\end{code}

The desugared version of that function follows below. It violates (naive?)
linearity by using $x$ twice, once to force the value and a second time to call
$f$. However, the $x$ passed as an argument is actually the case binder and it
must be handled in its own way. The case binder is the key (as seen
in~\ref{casebinder}) to solving this
and many other examples.
\begin{code}
($!)
  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
            (q :: Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: RuntimeRep))
      (@a)
      (@(b :: TYPE rep))
      (@(p :: Multiplicity))
      (@(q :: Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of x { __DEFAULT -> f x }
\end{code}

@@

\bibliographystyle{plain}
\bibliography{references}

>>>>

Since the publication of Linear Haskell~\cite{} and its release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types to a mainstream, pure, and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once (its arguments have multiplicity of one), for some
definition of \emph{consume} that I should revise here.

@@

\subsection{Handling the case binder}

@@

The definition for $\$!$ in \textbf{linear-base}\cite{} fails to lint after the
desugarer (before any optimisation takes place) with \emph{Linearity failure in
lambda: x 'Many $\not\subseteq$ p}
\begin{code}
($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
($!) f !x = f x
\end{code}

The desugared version of that function follows below. It violates (naive?)
linearity by using $x$ twice, once to force the value and a second time to call
$f$. However, the $x$ passed as an argument is actually the case binder and it
must be handled in its own way. The case binder is the key (as seen in ~\ref{caseBinderKey}) to solving this
and many other examples.
\begin{code}
($!)
  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
            (q :: Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: RuntimeRep))
      (@a)
      (@(b :: TYPE rep))
      (@(p :: Multiplicity))
      (@(q :: Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of x { __DEFAULT -> f x }
\end{code}

@@
